---
title: "fastFGEE: Functional Generalized Estimating Equations"
author: "Gabriel Loewinger"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{fastFGEE: Functional Generalized Estimating Equations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Suggested package location: vignettes/fastFGEE.Rmd -->
# Installation

The development version of `fastFGEE` can be installed with:

```{r, eval = FALSE}
remotes::install_github("gloewing/fastFGEE", build_vignettes = TRUE)
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", warning = FALSE, message = FALSE)
library(fastFGEE)
```

## Numerical dependencies

The default correlation solver handles regular-grid AR(1) and exchangeable
working correlation directly. `SuperGauss` is suggested rather than required
and is used only when `corr.solver = "supergauss"` is requested explicitly:

```{r, eval = FALSE}
install.packages("SuperGauss")
```

Irregularly sampled continuous-time AR(1) correlations use an exact
tridiagonal Markov precision implemented inside `fastFGEE` from the formulas of
Allevius (2018). Symmetric positive-definite coefficient-space solves use
registered Rcpp/LAPACK routines. The archived `irregulAR1` and `sanic` packages
are no longer dependencies.

# Introduction

`fastFGEE` fits functional generalized estimating equations for longitudinal
functional outcomes. The public workflow is:

1. fit or accept an initial function-on-scalar regression from `refund::pffr()`,
2. estimate a working correlation from the initial residuals,
3. choose smoothing parameters by fast cluster cross-validation, and
4. perform one penalized fGEE update followed by robust inference.

The public interface exposes only the validated one-step estimator. Historical
exact-GLS, pffr-only, legacy-engine, and fully iterated implementations remain
unexported for regression testing and future method development. Arguments such
as `exact`, `gee.fit`, `max.iter`, `tune.method`, and `working.engine` are not
accepted by `fgee()`.

The working-correlation directions are specified separately: `corr_long` acts
across repeated observations within a cluster and `corr_fn` acts along the
functional domain. The older `cov.type` interface is replaced by these two
arguments, and the older `sandwich` argument is replaced by `var.type`.

# Loading example data

The package includes a simulated longitudinal functional dataset named `d`.
The response is stored as an `AsIs` matrix column on a common functional grid:

```{r, eval = FALSE}
data("d", package = "fastFGEE")
dat <- d

head(as.matrix(dat$Y)[, 1:4])
head(dat[c("ID", "X1", "X2", "time")])
```

Each row is one longitudinal observation. `ID` identifies the independent
cluster, `time` orders repeated observations, and `Y` contains the functional
outcome evaluated on its grid.

# A basic one-step fit

The simplest current pattern is to specify `corr_long` and `corr_fn` directly.
For example, the following fits a one-step fGEE with AR1 correlation in the longitudinal direction and independence along the functional domain:

```{r, eval = FALSE}
fit_1step <- fgee(
  formula = Y ~ X1 + X2,
  data = dat,
  cluster = "ID",
  family = "binomial",
  time = "time",
  corr_long = "ar1",
  corr_fn = "independence",
  rho.smooth = TRUE,
  cv = "fastkfold",
  joint.CI = "wild",
  var.type = "sandwich"
)

fgee.plot(fit_1step)
```

The plotting method uses the fitted confidence intervals returned by `fgee()` and produces coefficient-function plots for the intercept and covariate effects.

# The supported one-step fit

Every public `fgee()` call performs one coefficient update. This is the fast
estimator described in the paper. There is no user switch between one-step,
exact, pffr-only, or fully iterated estimators in this development version.

```{r, eval = FALSE}
fit_1step <- fgee(
  formula = Y ~ X1 + X2,
  data = dat,
  cluster = "ID",
  family = "binomial",
  time = "time",
  corr_long = "ar1",
  corr_fn = "ar1",
  cv = "fastkfold",
  joint.CI = "wild",
  var.type = "sandwich"
)
```

A pre-fitted `refund::pffr()` model can still be supplied through `pffr.mod`;
that changes only the initial estimator, not the one-step nature of the public
fit. Internal historical paths are retained so package tests can continue to
compare the optimized engine with the original implementation.

# Smoothing-parameter selection

By default, and for every fastK selector, `fgee()` chooses smoothing parameters
by minimising the exact fast $K$-fold cross-validation criterion. The
`sp.method` argument selects the search strategy. `"sandwich_qreml"` is a
separate explicit opt-in working restricted quasi-likelihood criterion rather
than a cross-validation method; `"auto"` does not select it.

## What `"auto"` does

For Gaussian identity-link models, `"auto"` uses `"fastk_staged"`. The exact
held-out Gaussian loss can be written in terms of per-fold coefficient-space
sufficient statistics, so once those are built the objective no longer depends
on the number of functional observations. Tuning takes about the same time
whether the data have $10^5$ or $2 \times 10^7$ scalar observations.

For supported non-Gaussian families -- binomial with a logit link, Poisson,
Gamma and negative binomial with a log link, and beta regression with a logit
link -- `"auto"` uses `"fastk_grad_fast"`. There is no finite
sufficient-statistic reduction for the exact non-Gaussian held-out loss, so
every objective evaluation touches the data and the cost of tuning is dominated
by how many evaluations the search needs.

## Negative binomial and beta need their nuisance parameter

Gamma is the easy case: its dispersion enters the log-likelihood
multiplicatively, plus terms free of the linear predictor, so it cancels out of
the criterion and Gamma tuning never needed it. Negative binomial and beta are
different -- theta appears as $\theta(\mu - y)/(\mu + \theta)$ and the
precision inside $\psi(\mu\phi)$ -- so the criterion cannot be evaluated
without a value.

The value used is the one that formed the working variance, not a fresh
estimate. That is a correctness requirement rather than a convenience: the
held-out loss scores per-fold coefficients that were themselves computed from
$\bar W$ and $\bar d$ under that variance, so evaluating them under a
different nuisance would no longer be a held-out predictive loss for the model
being fitted.

Practically, that means supplying it through the family so it survives into the
initial fit:

```{r, eval = FALSE}
fit_nb <- fgee(
  formula = Y ~ X1 + X2, data = dat, cluster = "ID",
  family = mgcv::nb(),                 # or mgcv::nb(theta = 3) to fix it
  time = "time", corr_long = "exchangeable", corr_fn = "ar1"
)

fit_beta <- fgee(
  formula = Y ~ X1 + X2, data = dat, cluster = "ID",
  family = mgcv::betar(),
  time = "time", corr_long = "exchangeable", corr_fn = "ar1"
)
```

`MASS::negative.binomial()` will not work: `mgcv` rejects a plain family object
that lacks variance derivatives, inside `refund::pffr()`, before any fastFGEE
code runs. `mgcv::nb(theta = )` is the way to fix theta.

One caveat to carry. Whatever `mgcv::nb()` or `mgcv::betar()` reports is
estimated by REML under an **independence** likelihood, which is misspecified
for correlated functional data, and a flexible smoother can absorb variability
that would otherwise appear as overdispersion. The package therefore records
which nuisance value was used and updates it only at the designated final
working-state calculation; it does not re-estimate the nuisance parameter for
every candidate smoothing vector.

Dedicated selector simulations support retaining fastK as the non-Gaussian
default. Relative to `fastk_grad_fast`, fastK was never materially worse in the
examined tiers, whereas pure `sandwich_qreml` had coefficient-RMSE ratios as
large as 1.5653 for a concurrent negative-binomial functional-covariate design
and 1.4785 for the corresponding beta design under functional-independence
working-correlation misspecification. `qreml_fastk` reached the same fastK
answers but was 1.16--1.31 times slower, and `fastk_staged` showed no advantage
off Gaussian identity-link models. These results motivate the current automatic
dispatch; they do not imply that fastK is uniformly optimal outside the
validated simulation regimes.

## Why the non-Gaussian default changed in 0.3.0

Before 0.3.0 the non-Gaussian default was `"fastk_grad"`, which began by
scoring a dense 55-point common-scale ray, a qREML candidate and two axis
probes per smoothing parameter, and then ran four independent L-BFGS-B
optimisations. With three smoothing parameters that is roughly 62 exact
objective evaluations spent on choosing a starting point, against 4 to 46 spent
in the optimiser. The start search, not the optimisation, was the expense.

`"fastk_grad_fast"` optimises the *same* criterion from a cheaper start:

1. a coarse common-scale ray, $\lambda = 10^{c}\lambda_{\text{fREML}}$ for
   $c \in \{-3, \dots, 3\}$, extended outward while its best point sits on an
   endpoint and then refined by one-dimensional search;
2. a sparse anisotropic stage around that point, letting the smoothing
   parameters separate -- a $3^q$ grid for small $q$ and a coordinate sweep
   otherwise, so the cost grows linearly rather than exponentially in $q$;
3. a single continuous polish, with a second start only when the optimiser
   fails to converge, stops on a boundary, leaves a large projected gradient, or
   a runner-up candidate is close in criterion but far away in $\log\lambda$.

Probe evaluations drop from about 62 to about 34, isolated tuning is roughly
three times faster, and complete fits are 1.2 to 1.5 times faster, with the
larger gains at larger problems.

`"fastk_grad"` is still available and unchanged:

```{r, eval = FALSE}
fit_prev <- fgee(
  formula = Y ~ X1 + X2, data = dat, cluster = "ID",
  family = binomial(), time = "time",
  corr_long = "exchangeable", corr_fn = "ar1",
  sp.method = "fastk_grad"
)
```

## Evidence for the change

Changing a default selector changes which $\lambda$ is returned, so the two
selectors were compared in simulation rather than on timing alone.

Data were generated through a Gaussian copula: correlated latent normals are
mapped to uniforms and then through the inverse marginal distribution
function, so the response is *exactly* Bernoulli, Poisson or Gamma at the
intended mean while remaining correlated within a cluster. The marginal
coefficient functions that a GEE targets are therefore known in closed form.
This matters: adding a correlated error to the linear predictor instead would
make the marginal mean an attenuated version of the coefficient functions under
a non-linear link, and comparing estimates against them would show bias that is
an artefact of the simulation rather than a property of the estimator.

With 300 replicates per configuration, four cluster and visit configurations,
and both selectors seeing identical data at each replicate:

* no comparison of slope root-mean-square error was significant after
  adjustment for multiplicity;
* pointwise and simultaneous coverage of the studentised wild cluster bootstrap
  intervals differed by at most 0.008, against a Monte Carlo standard error of
  0.013;
* the same held under three working-correlation misspecification scenarios:
  independent longitudinally, independent functionally, and both.

## A caution about reading $\lambda$

The fastK criterion is very flat in places. Two selectors can return smoothing
parameters differing by orders of magnitude while agreeing on the fitted
coefficient functions to five significant figures. When comparing selectors,
compare the criterion value, the fitted coefficients, the effective degrees of
freedom, or coefficient error -- not $\lambda$.

## The optional compiled kernel

Each non-Gaussian objective evaluation computes, for every fold, a linear
predictor, a loss and its derivative, and a crossproduct for the score
contribution. In R that allocates several vectors as long as the fold for every
fold of every evaluation. A compiled kernel does the same arithmetic in one
pass with no intermediate allocation, making objective evaluations two to three
times faster.

The kernel is an accelerator, not a change of method: it reproduces the R
objective to about $10^{-16}$ in both the score and the gradient, so turning it
off cannot move a fitted model. It is used automatically when available, and
every entry point falls back to the R implementation otherwise. To disable it:

```{r, eval = FALSE}
fit_no_kernel <- fgee(
  formula = Y ~ X1 + X2, data = dat, cluster = "ID",
  family = binomial(), time = "time",
  corr_long = "exchangeable", corr_fn = "ar1",
  fastk.kernel = FALSE
)

# or for the whole session
options(fastFGEE.kernel = FALSE)
```

## The compiled working-correlation inverse

A second compiled kernel sits under `getD()`/`getW()`. When both correlations
are independence, exchangeable, or regular-grid AR(1), the closed-form
precision operators are applied in one allocation and two strided passes,
rather than by reshaping each cluster block, permuting it with `aperm()` and
permuting it back. On the largest benchmark cell that took the working-statistics
build from 2.27 s to 1.07 s and its share of a whole fit from 24% to 13%.

The same caveat as the fastK kernel applies with one addition. The kernel
reproduces the R operator to about $10^{-16}$, so it cannot meaningfully change
a fitted model — but the fastK criterion is flat in some directions, and a
perturbation that small can move a selected `lambda` on a plateau by orders of
magnitude. Compare `fit$beta` rather than `fit$lambda` when checking a run with
the kernel on against one with it off:

```{r, eval = FALSE}
options(fastFGEE.corr.kernel = FALSE)
```

FPCA functional correlation and an explicitly requested
`corr.solver = "supergauss"` remain separate paths. Irregular-grid AR(1) now
uses the package's compiled exact tridiagonal Markov precision, avoiding both a
dense inverse and the archived `irregulAR1` dependency.

## Workspace memory for non-Gaussian tuning

`fastk.memory = "balanced"` is the recommended default. It is the only current
layout that builds `prep$X_eval`, which is required by the compiled fastK
loss/gradient kernel. The `"speed"` and `"lowmem"` layouts therefore use the R
evaluation path.

Fresh-process measurements found no regime in which `"lowmem"` reduced peak
resident memory; it retained the long working table and was substantially
slower. It is deprecated and will be removed in a future release. `"speed"`
remains available for cases where a modest reduction in retained workspace is
worth slower R evaluation.

Peak memory must be assessed in a fresh R process because the long data, model
matrix, package dependencies, and transient working-statistics construction
usually dominate the retained fastK workspace.

# Working correlation structures

The new interface makes the two correlation directions explicit.
The main combinations are summarized below.

| `corr_long` | `corr_fn` | Working covariance interpretation |
|---|---|---|
| `"ar1"` or `"exchangeable"` | `"independence"` | Block diagonal in the **longitudinal** direction |
| `"independence"` | `"ar1"` or `"exchangeable"` | Block diagonal in the **functional** direction |
| `"independence"` | `"fpca"` | Block diagonal in the functional direction, with FPCA-based functional covariance |
| non-independence | non-independence | Separable / Kronecker product working covariance |
| `"ar1"` or `"exchangeable"` | `"fpca"` | Separable / Kronecker working covariance with a longitudinal parametric factor and an FPCA-based functional factor |

## Block diagonal versus Kronecker product working covariance

When exactly one direction is non-independent, the working covariance is **block diagonal**.
This means the package models dependence in only one direction:

- `corr_long != "independence"` and `corr_fn = "independence"` gives a block diagonal covariance across longitudinal observations,
- `corr_long = "independence"` and `corr_fn != "independence"` gives a block diagonal covariance across functional observations.

For the parametric block-diagonal cases (`"ar1"` or `"exchangeable"`), the package estimates a different correlation parameter for each block across the other dimension:

- longitudinal block-diagonal fits estimate `rho_long(s)` across the functional domain,
- functional block-diagonal fits estimate `rho_fn(t)` across longitudinal time.

If `rho.smooth = TRUE`, those block-specific correlation curves are smoothed across the corresponding index.

When **both** directions are non-independent, the package uses a **separable / Kronecker product** working covariance. In this case the correlation parameters are pooled and fixed within direction:

- one pooled longitudinal correlation parameter (or longitudinal factor), and
- one pooled functional correlation parameter (or functional factor).

So the main difference is:

- **block diagonal**: correlation varies by block,
- **Kronecker / separable**: correlation is fixed within direction and shared across blocks.

For Kronecker fits, each cluster must be observed on a complete longitudinal-by-functional grid.

## Longitudinal block-diagonal example

```{r, eval = FALSE}
fit_long_block <- fgee(
  formula = Y ~ X1 + X2,
  data = dat,
  cluster = "ID",
  family = "binomial",
  time = "time",
  corr_long = "ar1",
  corr_fn = "independence",
  rho.smooth = TRUE,
  joint.CI = "wild",
  var.type = "sandwich"
)
```

This fit models within-cluster dependence across repeated observations, and allows the longitudinal correlation estimate to vary over the functional domain.

## Functional block-diagonal example

```{r, eval = FALSE}
fit_fn_block <- fgee(
  formula = Y ~ X1 + X2,
  data = dat,
  cluster = "ID",
  family = "binomial",
  time = "time",
  corr_long = "independence",
  corr_fn = "exchangeable",
  rho.smooth = TRUE,
  joint.CI = "wild",
  var.type = "sandwich"
)
```

This fit models dependence along the functional domain while treating repeated observations as working independent.

## Separable / Kronecker example

```{r, eval = FALSE}
fit_sep <- fgee(
  formula = Y ~ X1 + X2,
  data = dat,
  cluster = "ID",
  family = "binomial",
  time = "time",
  corr_long = "ar1",
  corr_fn = "ar1",
  joint.CI = "wild",
  var.type = "sandwich"
)
```

This fit uses a separable working covariance with one longitudinal AR1 factor and one functional AR1 factor.

## FPCA-based functional covariance

`fastFGEE` also allows an FPCA-based covariance in the functional direction by setting `corr_fn = "fpca"`.

### FPCA functional block-diagonal fit

```{r, eval = FALSE}
fit_fpca_block <- fgee(
  formula = Y ~ X1 + X2,
  data = dat,
  cluster = "ID",
  family = "binomial",
  time = "time",
  corr_long = "independence",
  corr_fn = "fpca",
  joint.CI = "wild",
  var.type = "sandwich"
)
```

Here the longitudinal direction is working independent, and the functional blocks are modeled with an FPCA-based covariance estimate.

### FPCA plus longitudinal correlation: separable / Kronecker fit

```{r, eval = FALSE}
fit_fpca_sep <- fgee(
  formula = Y ~ X1 + X2,
  data = dat,
  cluster = "ID",
  family = "binomial",
  time = "time",
  corr_long = "ar1",
  corr_fn = "fpca",
  joint.CI = "wild",
  var.type = "sandwich"
)
```

This combines a parametric longitudinal factor with an FPCA-based functional factor in a separable working covariance.

# Supplying a pre-fit `refund::pffr()` object

A useful feature of the package is that you can fit `refund::pffr()` yourself and then pass the fitted object through `pffr.mod`.
This is particularly helpful when:

- the functional grid is irregular,
- you want to control the initial `pffr()` call directly, or
- you want to compare the initial fit with the fGEE update.

The main thing to remember is that `pffr.mod` supplies the aligned long-format response and design matrix, while `data` should still be the original wide data set used to define the cluster structure.

## Example: irregular functional grid

The following code shows the workflow using a user-supplied `pffr()` fit.

```{r, eval = FALSE}
# Start from the wide data object used above
Y_wide <- as.matrix(dat$Y)
colnames(Y_wide) <- paste0("Y_", seq_len(ncol(Y_wide)))

dat_wide <- data.frame(
  ID = dat$ID,
  X1 = dat$X1,
  X2 = dat$X2,
  time = dat$time,
  Y_wide
)

# Convert the matrix outcome to long format
# (shown here with tidyr for readability)
dat_long <- tidyr::pivot_longer(
  dat_wide,
  cols = tidyselect::starts_with("Y_"),
  names_to = "yindex",
  names_prefix = "Y_",
  values_to = "Y",
  values_drop_na = FALSE
)

dat_long$yindex <- as.integer(dat_long$yindex)
dat_long$time <- as.numeric(dat_long$time)

# Construct the ydata object expected by refund::pffr()
Y.mat <- data.frame(
  .obs = seq_len(nrow(dat_long)),
  .index = dat_long$yindex,
  .value = dat_long$Y
)

fit_pffr <- refund::pffr(
  formula = Y ~ X1 + X2,
  algorithm = "bam",
  family = binomial(),
  discrete = TRUE,
  yind = Y.mat$.index,
  ydata = Y.mat,
  bs.yindex = list(bs = "bs", k = 11),
  data = dat_long
)
```

Now update that initial fit with `fgee()`.
Notice that `pffr.mod` receives the fitted `pffr()` object, but `data` is still the original wide data frame.

```{r, eval = FALSE}
fit_from_pffr <- fgee(
  formula = Y ~ X1 + X2,
  pffr.mod = fit_pffr,
  data = dat,
  cluster = "ID",
  family = "binomial",
  time = "time",
  corr_long = "exchangeable",
  corr_fn = "independence",
  joint.CI = "wild",
  var.type = "sandwich"
)

fgee.plot(fit_from_pffr)
```

# Variance estimators and confidence intervals

`var.type` selects the coefficient covariance estimator. The historically named
`joint.CI` argument selects pointwise calibration and, when requested, a
whole-curve maximum-statistic simultaneous band. These are distinct targets:
pointwise coverage at each location does not imply simultaneous coverage over
the entire function.

```{r, eval = FALSE}
# Sandwich covariance with studentized wild-cluster calibration
fit_sw <- fgee(
  formula = Y ~ X1 + X2,
  data = dat,
  cluster = "ID",
  family = "binomial",
  time = "time",
  corr_long = "ar1",
  corr_fn = "ar1",
  var.type = "sandwich",
  joint.CI = "wild"
)

# Fast cluster-bootstrap covariance with the same wild calibration
fit_fb <- fgee(
  formula = Y ~ X1 + X2,
  data = dat,
  cluster = "ID",
  family = "binomial",
  time = "time",
  corr_long = "ar1",
  corr_fn = "ar1",
  var.type = "fastboot",
  boot.samps = 2000,
  joint.CI = "wild"
)
```

A practical default is `var.type = "sandwich"` with `joint.CI = "wild"`. The
resulting simultaneous band is a resampling-based whole-curve procedure, not a
guarantee of nominal finite-sample coverage. Reliability can deteriorate with
few independent clusters, concentrated cluster leverage, unstable nuisance
parameters, or unstable working-correlation estimates. In particular, making
fully iterated or exact experimental code inaccessible does not remove these
small-sample limitations.

# Public-interface notes

Compared with older versions:

- `cov.type` is replaced by `corr_long` and `corr_fn`;
- `sandwich` is replaced by `var.type`;
- public fitting is always one-step;
- `pffr.mod` remains the supported way to supply an initial `refund::pffr()` fit;
- experimental exact, legacy, pffr-only, and fully iterated controls are internal.

# Optimized one-step engine

The package uses a compact one-pass working-statistics engine by default for public one-step workflows. For cluster `i`, it forms

\[
W_i = Z_i^T R_i^{-1} Z_i, \qquad
 d_i = Z_i^T R_i^{-1} r_i,
\]

by applying the working-correlation inverse once to the augmented matrix
`cbind(Z_i, r_i)`. Regular AR(1) and exchangeable structures use direct
precision operators. The established `SuperGauss` Toeplitz implementation is
still available and can be forced with `corr.solver = "supergauss"`.

The public default remains the studentized wild-cluster procedure for pointwise intervals and optional simultaneous bands. Consequently,
`working.retain = "auto"` resolves to `"scores"`: the fit retains the compact
`p` by `N` matrix of cluster score contributions and aggregate bread matrices,
but not one `p` by `p` bread matrix for every cluster.

```{r, eval = FALSE}
fit <- fgee(
  Y ~ X1 + X2,
  data = dat,
  cluster = "ID",
  family = binomial(link = "logit"),
  time = "time",
  corr_long = "exchangeable",
  corr_fn = "ar1",
  sp.method = "auto",
  working.retain = "auto",
  corr.solver = "auto"
)
```

Available smoothing selectors include:

- `fastk_staged`: the published staged fast cluster-CV criterion;
- `fastk_grad`: continuous analytic-gradient optimization of the same exact
  fastK criterion;
- `sandwich_qreml`: an experimental coefficient-space working restricted
  quasi-likelihood with one-sided sandwich information scaling;
- `qreml_fastk`: qREML initialization followed by exact fastK optimization.

`qreml_fastk` is retained for reproducibility, but in the selector
simulations it reached the same fastK answers as `fastk_grad_fast` while taking
1.16--1.31 times as long. `fastk_staged` remains the Gaussian identity default;
it showed no advantage for the non-Gaussian families. Pure `sandwich_qreml` is
an opt-in experimental selector: it can be efficient when the working
correlation is well specified, but its worst examined coefficient-RMSE ratios
relative to `fastk_grad_fast` were 1.5653 for concurrent negative-binomial data
and 1.4785 for concurrent beta data under functional-independence
misspecification. It is therefore not the automatic default.

For Gaussian identity-link models, both staged and gradient fastK use fold
sufficient statistics after preparation. The tuning objective therefore does
not revisit the observation-level functional design at every candidate
smoothing vector.

Large fitted objects can be reduced after fitting with:

```{r, eval = FALSE}
fit_small <- fgee(
  Y ~ X1 + X2,
  data = dat,
  cluster = "ID",
  family = gaussian(),
  time = "time",
  joint.CI = FALSE,
  sp.method = "sandwich_qreml",
  keep.data = FALSE,
  keep.initial.fit = FALSE,
  keep.working.stats = FALSE
)
```

The legacy engine remains unexported and is reached only by package regression
tests and method-development code.
