---
title: "fastFGEE: Functional Generalized Estimating Equations"
author: "Gabriel Loewinger"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{fastFGEE: Functional Generalized Estimating Equations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Suggested package location: vignettes/fastFGEE.Rmd -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", warning = FALSE, message = FALSE)
library(fastFGEE)
```

# Introduction

`fastFGEE` fits functional generalized estimating equations (fGEE) for longitudinal functional outcomes.
The typical workflow is:

1. fit an initial function-on-scalar regression with `refund::pffr()`,
2. choose smoothing parameters with cluster cross-validation, and
3. optionally update the initial fit with one or more fGEE iterations.

The current interface uses separate arguments for the longitudinal and functional working correlations:

- `corr_long` controls the correlation structure across repeated observations within cluster,
- `corr_fn` controls the correlation structure along the functional domain.

This replaces the older single-argument `cov.type` interface used in the original vignette.
Likewise, the old `sandwich` argument is now replaced by `var.type`.

A useful way to think about the main fitting options is:

- `max.iter = 1`: one-step fGEE,
- `max.iter > 1`: fully-iterated final fit,
- `gee.fit = FALSE`: do **not** perform the GEE update; instead, keep the initial `pffr()` fit and compute robust inference for that fit.

For most users, `max.iter = 1` is the default fast estimator, while `max.iter = 100` is a practical way to approximate a fully-iterated fGEE fit.

# Loading example data

The source package currently includes a simulated example file named `binary_ar1_data`.
A robust way to locate it is:

```{r, eval = FALSE}
dat_path <- system.file("data", "binary_ar1_data", package = "fastFGEE")
if (dat_path == "") {
  dat_path <- "binary_ar1_data"
}
dat0 <- readRDS(dat_path)
```

The outcome is stored as a matrix inside a data frame column. A standard wide-format data object for `fgee()` looks like this:

```{r, eval = FALSE}
dat <- data.frame(
  Y = I(dat0$Y),
  X1 = dat0$X1,
  X2 = dat0$X2,
  ID = dat0$ID,
  time = dat0$time
)

head(as.matrix(dat$Y)[, 1:4])
head(dat[c("ID", "X1", "X2", "time")])
```

Here each row corresponds to one longitudinal observation, and `Y` is the functional outcome observed on a grid of length `L`.

# A basic one-step fit

The simplest current pattern is to specify `corr_long` and `corr_fn` directly.
For example, the following fits a one-step fGEE with AR1 correlation in the longitudinal direction and independence along the functional domain:

```{r, eval = FALSE}
fit_1step <- fgee(
  formula = Y ~ X1 + X2,
  data = dat,
  cluster = "ID",
  family = "binomial",
  time = "time",
  corr_long = "ar1",
  corr_fn = "independence",
  rho.smooth = TRUE,
  max.iter = 1,
  cv = "fastkfold",
  joint.CI = "wild",
  var.type = "sandwich"
)

fgee.plot(fit_1step)
```

The plotting method uses the fitted confidence intervals returned by `fgee()` and produces coefficient-function plots for the intercept and covariate effects.

# One-step, fully-iterated, and pffr-only fits

## One-step fGEE

A one-step fit is obtained with `max.iter = 1`.
This is the main fast estimator described in the paper.

```{r, eval = FALSE}
fit_1step <- fgee(
  formula = Y ~ X1 + X2,
  data = dat,
  cluster = "ID",
  family = "binomial",
  time = "time",
  corr_long = "ar1",
  corr_fn = "ar1",
  max.iter = 1,
  cv = "fastkfold",
  joint.CI = "wild",
  var.type = "sandwich"
)
```

## Fully-iterated final fit

To iterate the estimating equation update beyond one step, increase `max.iter`.
In the next example the smoothing parameters are still tuned using the default fast one-step tuning, but the final coefficient estimate is iterated until convergence or until the iteration cap is reached.

```{r, eval = FALSE}
fit_full <- fgee(
  formula = Y ~ X1 + X2,
  data = dat,
  cluster = "ID",
  family = "binomial",
  time = "time",
  corr_long = "ar1",
  corr_fn = "ar1",
  max.iter = 100,
  cv = "fastkfold",
  joint.CI = "wild",
  var.type = "sandwich"
)

fit_full$n_iter
fit_full$converged
```

## Fully-iterated tuning and fully-iterated final fit

If you also want the cross-validation stage to use the fully-iterated fit instead of the one-step approximation, set `tune.method = "fully-iterated"`.

```{r, eval = FALSE}
fit_full_tuned <- fgee(
  formula = Y ~ X1 + X2,
  data = dat,
  cluster = "ID",
  family = "binomial",
  time = "time",
  corr_long = "ar1",
  corr_fn = "ar1",
  max.iter = 100,
  tune.method = "fully-iterated",
  cv = "fastkfold",
  joint.CI = "wild",
  var.type = "sandwich"
)
```

## `gee.fit = FALSE`: robust inference for the initial `pffr()` fit only

Set `gee.fit = FALSE` when you want to keep the initial `pffr()` mean fit and compute robust inference without performing the GEE update.
This is useful when you want a direct comparison between:

- the initial `pffr()` fit,
- the one-step fGEE fit, and
- the fully-iterated fGEE fit.

```{r, eval = FALSE}
fit_pffr_robust <- fgee(
  formula = Y ~ X1 + X2,
  data = dat,
  cluster = "ID",
  family = "binomial",
  time = "time",
  gee.fit = FALSE,
  max.iter = 1,
  joint.CI = "wild",
  var.type = "sandwich"
)
```

When `gee.fit = FALSE`, the package does not construct a non-independence working covariance for the update step. In practice this is the cleanest way to obtain **robust sandwich-based inference for the initial `pffr()` fit only**.

A direct comparison of the three fits can be made as follows:

```{r, eval = FALSE}
p_pffr <- fgee.plot(
  fit_pffr_robust,
  xlab = "Functional Domain",
  title_names = c("pffr: Intercept", "pffr: X1", "pffr: X2")
)

p_1step <- fgee.plot(
  fit_1step,
  xlab = "Functional Domain",
  title_names = c("One-step: Intercept", "One-step: X1", "One-step: X2")
)

p_full <- fgee.plot(
  fit_full,
  xlab = "Functional Domain",
  title_names = c("Fully-iterated: Intercept", "Fully-iterated: X1", "Fully-iterated: X2")
)

gridExtra::grid.arrange(p_pffr, p_1step, p_full, nrow = 3)
```

# Working correlation structures

The new interface makes the two correlation directions explicit.
The main combinations are summarized below.

| `corr_long` | `corr_fn` | Working covariance interpretation |
|---|---|---|
| `"ar1"` or `"exchangeable"` | `"independence"` | Block diagonal in the **longitudinal** direction |
| `"independence"` | `"ar1"` or `"exchangeable"` | Block diagonal in the **functional** direction |
| `"independence"` | `"fpca"` | Block diagonal in the functional direction, with FPCA-based functional covariance |
| non-independence | non-independence | Separable / Kronecker product working covariance |
| `"ar1"` or `"exchangeable"` | `"fpca"` | Separable / Kronecker working covariance with a longitudinal parametric factor and an FPCA-based functional factor |

## Block diagonal versus Kronecker product working covariance

When exactly one direction is non-independent, the working covariance is **block diagonal**.
This means the package models dependence in only one direction:

- `corr_long != "independence"` and `corr_fn = "independence"` gives a block diagonal covariance across longitudinal observations,
- `corr_long = "independence"` and `corr_fn != "independence"` gives a block diagonal covariance across functional observations.

For the parametric block-diagonal cases (`"ar1"` or `"exchangeable"`), the package estimates a different correlation parameter for each block across the other dimension:

- longitudinal block-diagonal fits estimate `rho_long(s)` across the functional domain,
- functional block-diagonal fits estimate `rho_fn(t)` across longitudinal time.

If `rho.smooth = TRUE`, those block-specific correlation curves are smoothed across the corresponding index.

When **both** directions are non-independent, the package uses a **separable / Kronecker product** working covariance. In this case the correlation parameters are pooled and fixed within direction:

- one pooled longitudinal correlation parameter (or longitudinal factor), and
- one pooled functional correlation parameter (or functional factor).

So the main difference is:

- **block diagonal**: correlation varies by block,
- **Kronecker / separable**: correlation is fixed within direction and shared across blocks.

For Kronecker fits, each cluster must be observed on a complete longitudinal-by-functional grid.

## Longitudinal block-diagonal example

```{r, eval = FALSE}
fit_long_block <- fgee(
  formula = Y ~ X1 + X2,
  data = dat,
  cluster = "ID",
  family = "binomial",
  time = "time",
  corr_long = "ar1",
  corr_fn = "independence",
  rho.smooth = TRUE,
  max.iter = 1,
  joint.CI = "wild",
  var.type = "sandwich"
)
```

This fit models within-cluster dependence across repeated observations, and allows the longitudinal correlation estimate to vary over the functional domain.

## Functional block-diagonal example

```{r, eval = FALSE}
fit_fn_block <- fgee(
  formula = Y ~ X1 + X2,
  data = dat,
  cluster = "ID",
  family = "binomial",
  time = "time",
  corr_long = "independence",
  corr_fn = "exchangeable",
  rho.smooth = TRUE,
  max.iter = 1,
  joint.CI = "wild",
  var.type = "sandwich"
)
```

This fit models dependence along the functional domain while treating repeated observations as working independent.

## Separable / Kronecker example

```{r, eval = FALSE}
fit_sep <- fgee(
  formula = Y ~ X1 + X2,
  data = dat,
  cluster = "ID",
  family = "binomial",
  time = "time",
  corr_long = "ar1",
  corr_fn = "ar1",
  max.iter = 1,
  joint.CI = "wild",
  var.type = "sandwich"
)
```

This fit uses a separable working covariance with one longitudinal AR1 factor and one functional AR1 factor.

## FPCA-based functional covariance

`fastFGEE` also allows an FPCA-based covariance in the functional direction by setting `corr_fn = "fpca"`.

### FPCA functional block-diagonal fit

```{r, eval = FALSE}
fit_fpca_block <- fgee(
  formula = Y ~ X1 + X2,
  data = dat,
  cluster = "ID",
  family = "binomial",
  time = "time",
  corr_long = "independence",
  corr_fn = "fpca",
  max.iter = 1,
  joint.CI = "wild",
  var.type = "sandwich"
)
```

Here the longitudinal direction is working independent, and the functional blocks are modeled with an FPCA-based covariance estimate.

### FPCA plus longitudinal correlation: separable / Kronecker fit

```{r, eval = FALSE}
fit_fpca_sep <- fgee(
  formula = Y ~ X1 + X2,
  data = dat,
  cluster = "ID",
  family = "binomial",
  time = "time",
  corr_long = "ar1",
  corr_fn = "fpca",
  max.iter = 1,
  joint.CI = "wild",
  var.type = "sandwich"
)
```

This combines a parametric longitudinal factor with an FPCA-based functional factor in a separable working covariance.

# Supplying a pre-fit `refund::pffr()` object

A useful feature of the package is that you can fit `refund::pffr()` yourself and then pass the fitted object through `pffr.mod`.
This is particularly helpful when:

- the functional grid is irregular,
- you want to control the initial `pffr()` call directly, or
- you want to compare the initial fit with the fGEE update.

The main thing to remember is that `pffr.mod` supplies the aligned long-format response and design matrix, while `data` should still be the original wide data set used to define the cluster structure.

## Example: irregular functional grid

The following code shows the workflow using a user-supplied `pffr()` fit.

```{r, eval = FALSE}
# Start from the wide data object used above
Y_wide <- as.matrix(dat$Y)
colnames(Y_wide) <- paste0("Y_", seq_len(ncol(Y_wide)))

dat_wide <- data.frame(
  ID = dat$ID,
  X1 = dat$X1,
  X2 = dat$X2,
  time = dat$time,
  Y_wide
)

# Convert the matrix outcome to long format
# (shown here with tidyr for readability)
dat_long <- tidyr::pivot_longer(
  dat_wide,
  cols = tidyselect::starts_with("Y_"),
  names_to = "yindex",
  names_prefix = "Y_",
  values_to = "Y",
  values_drop_na = FALSE
)

dat_long$yindex <- as.integer(dat_long$yindex)
dat_long$time <- as.numeric(dat_long$time)

# Construct the ydata object expected by refund::pffr()
Y.mat <- data.frame(
  .obs = seq_len(nrow(dat_long)),
  .index = dat_long$yindex,
  .value = dat_long$Y
)

fit_pffr <- refund::pffr(
  formula = Y ~ X1 + X2,
  algorithm = "bam",
  family = binomial(),
  discrete = TRUE,
  yind = Y.mat$.index,
  ydata = Y.mat,
  bs.yindex = list(bs = "bs", k = 11),
  data = dat_long
)
```

Now update that initial fit with `fgee()`.
Notice that `pffr.mod` receives the fitted `pffr()` object, but `data` is still the original wide data frame.

```{r, eval = FALSE}
fit_from_pffr <- fgee(
  formula = Y ~ X1 + X2,
  pffr.mod = fit_pffr,
  data = dat,
  cluster = "ID",
  family = "binomial",
  time = "time",
  corr_long = "exchangeable",
  corr_fn = "independence",
  max.iter = 1,
  joint.CI = "wild",
  var.type = "sandwich"
)

fgee.plot(fit_from_pffr)
```

# Variance estimators and confidence intervals

The variance estimator and the joint confidence interval construction are controlled separately.

- `var.type` chooses the covariance estimator for the coefficient vector,
- `joint.CI` chooses how the joint critical values are obtained.

Common choices are:

```{r, eval = FALSE}
# Sandwich variance + wild-bootstrap critical values
fit_sw <- fgee(
  formula = Y ~ X1 + X2,
  data = dat,
  cluster = "ID",
  family = "binomial",
  time = "time",
  corr_long = "ar1",
  corr_fn = "ar1",
  var.type = "sandwich",
  joint.CI = "wild"
)

# Fast cluster bootstrap variance + wild-bootstrap critical values
fit_fb <- fgee(
  formula = Y ~ X1 + X2,
  data = dat,
  cluster = "ID",
  family = "binomial",
  time = "time",
  corr_long = "ar1",
  corr_fn = "ar1",
  var.type = "fastboot",
  boot.samps = 2000,
  joint.CI = "wild"
)

# Sandwich variance + parametric joint critical values
fit_param <- fgee(
  formula = Y ~ X1 + X2,
  data = dat,
  cluster = "ID",
  family = "binomial",
  time = "time",
  corr_long = "ar1",
  corr_fn = "ar1",
  var.type = "sandwich",
  joint.CI = "parametric"
)
```

A practical default for many applications is `var.type = "sandwich"` with `joint.CI = "wild"`.
This is also the most direct setting when comparing `gee.fit = FALSE`, one-step, and fully-iterated fits.

# Gaussian exact mode

For Gaussian responses with the identity link, `exact = TRUE` uses the exact penalized GLS-style update instead of the approximate one-step update.
For non-Gaussian families, `exact = TRUE` is ignored.

```{r, eval = FALSE}
fit_gaussian_exact <- fgee(
  formula = Y ~ X1 + X2,
  data = dat,
  cluster = "ID",
  family = "gaussian",
  time = "time",
  corr_long = "ar1",
  corr_fn = "ar1",
  exact = TRUE,
  max.iter = 1,
  joint.CI = "wild",
  var.type = "sandwich"
)
```

# Notes on the updated interface

Compared with the older vignette, the main argument changes are:

- `cov.type` is replaced by `corr_long` and `corr_fn`,
- `sandwich` is replaced by `var.type`,
- `max.iter` now explicitly controls one-step versus fully-iterated fitting,
- `gee.fit = FALSE` gives robust inference for the initial `pffr()` fit,
- `pffr.mod` is the supported way to pass in a pre-fit `refund::pffr()` object.

In most analyses, the current recommended pattern is:

```{r, eval = FALSE}
fit <- fgee(
  formula = Y ~ X1 + X2,
  data = dat,
  cluster = "ID",
  family = "binomial",
  time = "time",
  corr_long = "ar1",
  corr_fn = "ar1",
  max.iter = 1,
  cv = "fastkfold",
  joint.CI = "wild",
  var.type = "sandwich"
)
```
